---
title: 'COVID-19 guide design (Cas13a): RPA/LAMP amplicon overlap, minus strand'
author: "Amanda Mok"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, echo=F}
knitr::opts_chunk$set(echo=F, fig.width=12, fig.height=8, warning=F)

library(ggplot2)
library(patchwork)
library(prodlim)

source("~/covid-19/scripts/helper.R")
```

```{r load_data}
num_covid19_strains <- as.numeric(strsplit(system("wc -l ~/covid-19/ref_data/gisaid_cov2020_alignment.txt", intern=T),
                                           split=" ")[[1]][1])
num_human_CoV <- as.numeric(system(paste("grep ^'>' ~/covid-19/ref_data/human_CoV.fa | wc -l"), intern=T))

coverage_id <- "PRJNA616446"
abundance_binSize <- 300
abundance <- load_coverage(coverage_id, abundance_binSize)

windows <- read.table("cas13a_minusStrand_results_summary.txt", header=T, stringsAsFactors=F)
output_columns <- c("target", "spacer", "start", "antitag", "target_basepairing_prob", "match_against_bosTau9")
```

# Design

AIM: High quality spacers that overlap with RPA/LAMP amplicons

### 1. Pre-processing

- Remove spacers with GUUU antitags
- Remove spacers that interfere with crRNA hairpin
- Remove spacers with ≥4 internal base-pairs

```{r subset_1}
subset_1 <- subset(windows, antitag != "GUUU")
subset_1 <- subset(subset_1, has_crRNA_hairpin)
subset_1 <- subset(subset_1, crRNA_spacer_basepairs <= 4)
filter_1 <- "Preprocessed"
plot_diagnostic(subset_1, abundance, filter=filter_1,
                alpha=0.2, jitter_x=0.01, jitter_y=0.05,
                var1="sensitivity_0", var1_name="Sensitivity", var1_desc="% SARS-CoV-2 genomes captured",
                var2="specificity", var2_name="Specificity", var2_desc="% human coronaviruses missed",
                var3="match_against_hg38", var3_name="Reactivity: hg38", var3_desc="# alignments to human transcriptome")
```

### 2. Select spacers with high specificity

- No alignments to other human coronaviruses
- No alignments to human transcriptome

```{r subset_2}
subset_2 <- subset(subset_1, specificity==1 & match_against_hg38==0)
filter_2 <- paste(filter_1, "100% specificity", sep=", ")
plot_diagnostic(subset_2, abundance, filter=filter_2,
                alpha=0.2, jitter_x=0.01, jitter_y=0.05,
                var1="sensitivity_0", var1_name="Sensitivity", var1_desc="% SARS-CoV-2 genomes captured",
                var2="GC_content", var2_name="% GC", var2_desc="Spacer GC content",
                var3="target_basepairing_prob", var3_name="Protospacer structure", 
                var3_desc="Propensity for viral protospacer to base-pair")
```

### 3. Select spacers that overlap with RPA/LAMP amplicons

```{r subset_3}
# read in alignments
amplicons <- system("grep -v @ ~/covid-19/isothermal_amplification/amplicons_mapped.sam | cut -f 1,2,4,10", intern=T)
amplicons <- data.frame(matrix(unlist(strsplit(amplicons, split="\t")), byrow=T, ncol=4), stringsAsFactors=F)
colnames(amplicons) <- c("name", "flag", "position", "sequence")

# process name
amplicons$amplicon <- sapply(amplicons$name, function(x) strsplit(x, split=";")[[1]][1])
amplicons$FW <- sapply(amplicons$name, function(x) strsplit(strsplit(x, split=";")[[1]][2], split=":")[[1]][1])
amplicons$FW_seq <- sapply(amplicons$name, function(x) strsplit(strsplit(x, split=";")[[1]][2], split=":")[[1]][2])
amplicons$REV <- sapply(amplicons$name, function(x) strsplit(strsplit(x, split=";")[[1]][3], split=":")[[1]][1])
amplicons$REV_seq <- sapply(amplicons$name, function(x) strsplit(strsplit(x, split=";")[[1]][3], split=":")[[1]][2])
amplicons <- amplicons[,-1]

# find spacers that overlap with amplicons
subset_3 <- lapply(seq.int(nrow(amplicons)),
                           function(x) {
                             ## account for primer lengths, spacer length, and antitag
                             amplicon_start <- as.numeric(amplicons$position[x]) + nchar(amplicons$FW_seq[x])-1
                             amplicon_stop <- amplicon_start + nchar(amplicons$sequence[x]) - nchar(amplicons$REV_seq[x])+1
                             spacer_start <- amplicon_start + 20+4-1 # account for spacer length and antitag
                             selection <- subset(subset_2, start>=spacer_start & start<=amplicon_stop)
                             if(nrow(selection)>0) {
                               return(data.frame(selection,
                                                 amplicon=amplicons$amplicon[x],
                                                 FW=amplicons$FW[x],
                                                 FW_seq=amplicons$FW_seq[x],
                                                 REV=amplicons$REV[x],
                                                 REV_seq=amplicons$REV_seq[x]))
                             } else {
                               return(NULL)
                             }
                           })
subset_3 <- do.call(rbind, subset_3)

# output spacers
# write.table(subset_3, quote=F, row.names=F, sep="\t",
#             file=paste0("~/covid-19/isothermal_amplification/guides_amplicon_overlap_minusStrand", 
#                         format(Sys.time(), "%Y%m%d"), ".txt"))

# diagnostic plot
subset_3$amplicon_label <- gsub(":.*", "", subset_3$amplicon) 
genome_breaks <- seq(from=1, to=30000, by=300)
axis_scale <- max(abundance$coverage)/max(summary(cut(subset_3$start, breaks=genome_breaks)))
abundance$scaled_coverage <- abundance$coverage / axis_scale
plot_position <- ggplot() + 
  geom_histogram(data=subset_3, aes(start), binwidth=300, fill=2, col=2, alpha=0.5) + xlim(0, 30000) + 
  geom_area(data=abundance, aes(x=bin, y=scaled_coverage), fill=1, col=1, alpha=0.2) +
  scale_y_continuous(name="guide", sec.axis=sec_axis(~(axis_scale)*., name="virus")) +
  theme_bw() + ggtitle("Genomic position", subtitle=paste(length(unique(subset_3$start)), "spacers")) + 
  xlab("position") + ylab("# spacers") +
  theme(axis.title.y=element_text(color=1), axis.title.y.left = element_text(color=2))
plot_sensitivity <- ggplot(subset_3, aes(x=amplicon_label, y=sensitivity_0, col=amplicon_label)) + theme_bw() + 
  geom_jitter(alpha=0.5, size=2, height=0.005, width=0.4) + ggtitle("Sensitivity") + 
  xlab("") + ylab("") + theme(legend.position="none") + scale_x_discrete(labels=summary(as.factor(subset_3$amplicon_label)))
plot_count <- ggplot(subset_3, aes(amplicon_label, fill=amplicon_label)) + theme_bw() + geom_bar() +
  theme(axis.text.x=element_text(angle=90), legend.position="none") + xlab("")
plot_position / plot_sensitivity / plot_count + plot_layout(heights=c(1,2,1))
```